What to simulate? Inferring the direction for mental rotation

### 2014-01-31 man/cogsci-2014/cogsci-2014.tex
Misc changes to paper

\input{../../results/D/response_time.tex}
\input{../../results/D/response_time_corrs.tex}
\input{../../results/D/accuracy.tex}
\input{../../results/D/accuracy_corrs.tex}
\input{../../results/D/trial_time_corrs.tex}
\input{../../results/D/trial_accuracy_corrs.tex}
\input{../../results/D/num_chance.tex}
\input{../../results/D/theta_time_corrs.tex}
\input{../../results/D/theta_accuracy_corrs.tex}

@@

The original congruency hypothesis \cite{Shepard1971} actually is a
rational solution which acheives this goal, in the sense that the
smallest amount of computation coincides with rotating through the
minimum angle. However, it relies on assumptions of \textit{a priori}
knowledge (either of the minimum angle or the correspondence of points
between shapes). Noting that a rational solution need not necessarily
maintain a single trajectory of rotation, we explore an alternative
rational model, which--rather than assuming the angle of rotation
\textit{a priori}--engages in an ``active sampling'' strategy.

@@

One way to deal with the problem of choosing a threshold would simply
be to perform an exhaustive search or use a global optimisation
strategy. However, these strategies would not result in the linear
response time found by \citeA{Shepard1971}. A second alternative,
which could potentially demonstrate the linear response time result,
might instead just perform a Hill Climbing (\Hc{}) search for a
\textit{local} maximum. As with the \Oc{} and \Th{} models,
Eq.~\ref{eq:lh-ratio} becomes a maximum likelihood ratio. Unlike the
global maximum, however, it is not guaranteed that a local maximum of
the correct hypothesis will be greater than a local maximum of the
incorrect hypothesis.

@@

@@

We randomly generated 20 shapes of 5 or 6 vertices (e.g.,
Fig.~\ref{fig:model-traces}). For each shape, we computed 20 ``same''
and 20 ``flipped'' stimuli pairs, with rotations ($\theta$) spaced at
$20^\prime$ increments between 0 and 360 (with 0 and 180 degree
rotations repeated twice, in order to gather an equal number of
responses for each angle between 0 and 180). ``Same'' pairs were
created by rotating $\Xa$ by $\theta$; ``flipped'' pairs first
reflected $\Xa$ across the $y$-axis, then rotated by $\theta$.

@@

On each trial, particpants were instructed to press the `b' key to
begin and to focus on the fixation cross that appeared for 750
milliseconds afterwards. The two images were then presented
side-by-side, each at 300px $\times$ 300px, and participants could
press `s' to indicate they thought the images depicted the same
object, or `d' to indicate they thought the images depicted flipped
objects.

While there was no limit on response time, participants were urged to
repond as quickly as possible while still being
accurate. Specifically, we asked them to aim for at least 85\%
accuracy in the experimental blocks, and provided a counter to keep
them informed of their score.

@@

For analyses of response time, confidence intervals were computed
using a bootstrap analysis (sampling with replacement) with 1000
bootstrap samples. Response times were only only considered for
correct responses.  We also used a bootstrap analysis (with 10000
bootstrap samples) to compute the confidence intervals for both
Spearman ($r_s$) and Pearson ($r$) correlations.

@@

Out of the 247 participants, 200 (81\%) were included in our
analyses. Of the rest, we excluded 10 (4\%) because of an experimental
error, 6 (2.4\%) because they had already compelted a previous version
of the experiment, and 31 (12.6\%) because they failed a comprehension
check. A participant was marked as passing the comprehension check if
they answered at least 85\% of ``easy'' trials correctly. ``Easy''
trials were defined to be those for which the presented shapes
differed by a rotation of no more than 20 degrees. Additionally, we
excluded individual trials for which the response time exceeded 20
seconds.

For each model, we ran 50 samples for each of the same stimuli that
participants viewed (with the exception of stimuli with rotations of
$0^\prime$ or $180^\prime$, in which case we ran double the number of
samples). The step size parameter ($\sigma_\epsilon$) was
approximately fit to each of the models, resulting in
$\sigma_\epsilon=0.6$ for the \Bq{} model, and $\sigma_\epsilon=0.1$
for all other models. We also ran the models under two different
priors, $p(h=0)=0.5$ (the ``equal'' prior) and $p(h=0)=0.55$ (the
``unequal'' prior), though this only had an effect for the \Bq{}
model. \TODO{actually, did it affect HC?  check this}

\begin{figure}[t]
  \centering
  \subfloat{\includegraphics[width=0.24\textwidth]{../../figures/D/trial_accuracy.pdf}}%
  \subfloat{\includegraphics[width=0.24\textwidth]{../../figures/D/trial_time.pdf}}
  \caption{\textbf{Effect of learning.} There was a clear effect of
    learning on participants' responses, especially in the first half
    of the experiment. Left: accuracy as a function of trial. Right:
    response time as a function of trial.}
  \label{fig:learning}
\end{figure}

@@

The average response time across all stimuli was \ExpTime{} seconds,
and the average response time for an individual stimulus ranged from
\ExpTimeMin{} to \ExpTimeMax{} seconds.  As expected, the minimum
angle of rotation was significanty monotonically correlated with
average per-stimulus response times, both for flipped pairs
(\ExpThetaTimeCorrFlipped{}) and same pairs
(\ExpThetaTimeCorrSame{}). This was also the case for the number of
actions taken by the \Oc{} model (\OcThetaTimeCorr{}), the \Th{} model
(\ThThetaTimeCorr{}), and the \Bq{} model (\BqThetaTimeCorr{}), but
not the \Hc{} model (\HcThetaTimeCorr{}). The top row of
Fig.~\ref{fig:response-time-accuracy} shows the overall effect of
rotation on response times.

The number of actions taken by the \Oc{} model was the best predictor
of human response times, with a correlation of \ExpOcTimeCorr{}. The
\Th{} model \Bq{} model with $p(h=0)=0.55$ were the next best (for
\Th{}, \ExpThTimeCorr{}; for \Bq{}, \ExpBqpTimeCorr{}). Not far behind
was the BQ model with equal prior (\ExpBqTimeCorr{}). The \Hc{} model
was barely correlated with people (\ExpHcTimeCorr{}).

Across all stimuli, participants' response times for ``flipped'' pairs
were significantly higher than those for ``same'' pairs
\TODO{confidence intervals} \TODO{flipped response times higher than
  same response times for smaller angles?}

@@

The average accuracy across all stimuli was \ExpAccuracy{}, though
there were \ExpNumChance{} individual stimuli (out of 720) for which
people were not above chance. Of the models, the \Oc{} and \Th{}
models peformed perfectly, with 100\% accuracy. The \Bq{} model was
very accurate overall (equal prior: \BqAccuracy{}, unequal prior:
\BqpAccuracy{}). With the equal prior, there was \BqNumChance{}
stimulus for which it was not above chance; with the unequal prior,
there were \BqpNumChance{}. The \Hc{} model was not particularly
accurate, and although it did perform above chance overall
(\HcAccuracy{}), there were \HcNumChance{} individual stimuli for
which it was not above chance.

The minimum angle was also correlated with participants' average
per-stimulus accuracy: for flipped pairs,
\ExpThetaAccuracyCorrFlipped{}, and for same pairs,
\ExpThetaAccuracyCorrSame{}. There was also a significant effect for
the \Hc{} model (\HcThetaAccuracyCorr{}) and the \Bq{} model (equal
prior: \BqThetaAccuracyCorr{}, unequal prior:
\BqpThetaAccuracyCorr{}). The bottom row of
Fig.~\ref{fig:response-time-accuracy} shows the effect of angle on
accuracy for both participants and models.

There was a moderate correlation between the accuracy of the \Hc{}
model and human accuracy (\ExpHcAccuracyCorr{}). The \Bq{} model had a
similar correlation (equal prior: \ExpBqAccuracyCorr{}, unequal prior:
\ExpBqpAccuracyCorr{}). Correlations for the other models were
undefined, because they had 100\% accuracy.

\begin{figure*}[t]
  \begin{center}
    \includegraphics[width=\textwidth]{../../figures/D/response_time_accuracy.pdf}
    \caption{\textbf{Response time and accuracy comparison.} Top:
      response time of correct responses as a function of the minimum
      angle of rotation. Bottom: accuracy as a function of the minumum
      angle of rotation. All error bars are 95\% confidence
      intervals.}
    \label{fig:response-time-accuracy}
  \end{center}
\end{figure*}

@@

We replicated the previous monotonic response time result, though not
the linear resulte. Actually kind of similar to \cite{Gardony:2013gn}
\TODO{elaborate}

>>>>

\input{../../results/D/overall_response_time.tex}
\input{../../results/D/overall_accuracy.tex}
\input{../../results/D/num_chance.tex}
\input{../../results/D/trial_accuracy_corrs.tex}
\input{../../results/D/trial_time_corrs.tex}

@@

The original congruency hypothesis \cite{Shepard1971} actually is a
rational solution which achieves this goal, in the sense that the
smallest amount of computation coincides with rotating through the
minimum angle. However, it relies on assumptions of \textit{a priori}
knowledge (either of the minimum angle or the correspondence of points
between shapes). Noting that a rational solution need not necessarily
maintain a single trajectory of rotation, we explore an alternative
rational model, which--rather than assuming the angle of rotation
\textit{a priori}--engages in an ``active sampling'' strategy.

@@

One way to deal with the problem of choosing a threshold would simply
be to perform an exhaustive search or use a global optimization
strategy. However, these strategies would not result in the linear
response time found by \citeA{Shepard1971}. A second alternative,
which could potentially demonstrate the linear response time result,
might instead just perform a Hill Climbing (\Hc{}) search for a
\textit{local} maximum. As with the \Oc{} and \Th{} models,
Eq.~\ref{eq:lh-ratio} becomes a maximum likelihood ratio. Unlike the
global maximum, however, it is not guaranteed that a local maximum of
the correct hypothesis will be greater than a local maximum of the
incorrect hypothesis.

@@

\begin{figure}[t]
  \centering
  \subfloat{\includegraphics[width=0.24\textwidth]{../../figures/D/trial_accuracy.pdf}}%
  \subfloat{\includegraphics[width=0.24\textwidth]{../../figures/D/trial_time.pdf}}
  \caption{\textbf{Effect of learning.} There was a clear effect of
    learning on participants' responses, especially in the first half
    of the experiment. Left: accuracy as a function of trial. Right:
    response time in milliseconds as a function of trial.}
  \label{fig:learning}
\end{figure}

@@

We randomly generated 20 shapes of 5 or 6 vertices (e.g.,
Fig.~\ref{fig:model-traces}). For each shape, we computed 20 ``same''
and 20 ``flipped'' stimuli pairs, with rotations ($\theta$) spaced at
$20^\prime$ increments between 0 and 360 (with 0 and 180 degree
rotations repeated twice, in order to gather an equal number of
responses for each angle between 0 and 180). ``Same'' pairs were
created by rotating $\Xa$ by $\theta$; ``flipped'' pairs were first
reflected $\Xa$ across the $y$-axis, then rotated by $\theta$.

@@

On each trial, participants were instructed to press the `b' key to
begin and to focus on the fixation cross that appeared for 750
milliseconds afterwards. The two images were then presented
side-by-side, each at 300px $\times$ 300px, and participants could
press `s' to indicate they thought the images depicted the same
object, or `d' to indicate they thought the images depicted flipped
objects.

While there was no limit on response time, participants were urged to
respond as quickly as possible while still being
accurate. Specifically, we asked them to aim for at least 85\%
accuracy in the experimental blocks, and provided a counter to keep
them informed of their score.

@@

\begin{figure*}[t]
  \begin{center}
    \includegraphics[width=\textwidth]{../../figures/D/response_time_accuracy.pdf}
    \caption{\textbf{Response time and accuracy comparison.} Top:
      response time of correct responses as a function of the minimum
      angle of rotation. Bottom: accuracy as a function of the minimum
      angle of rotation. All error bars are 95\% confidence
      intervals.}
    \label{fig:response-time-accuracy}
  \end{center}
\end{figure*}

For analyses of response time, confidence intervals around harmonic
means were computed using a bootstrap analysis using 10000 bootstrap
samples (sampled with replacement). Response times were only
considered for correct responses.  We also used a bootstrap analysis
of 10000 bootstrap samples to compute the confidence intervals around
both Spearman ($r_s$) and Pearson ($r$) correlations.

@@

Out of the 247 participants, 200 (81\%) were included in our
analyses. Of the rest, we excluded 10 (4\%) because of an experimental
error, 6 (2.4\%) because they had already completed a previous version
of the experiment, and 31 (12.6\%) because they failed a comprehension
check. A participant was marked as passing the comprehension check if
they answered at least 85\% of ``easy'' trials correctly. ``Easy''
trials were defined to be those for which the presented shapes
differed by a rotation of no more than 20 degrees. Additionally, we
excluded 82 individual trials (out of 40000) for which the response
time was either less than 100 milliseconds or greater than 20 seconds.

For each model, we ran 50 samples for each stimulus that participants
viewed (with the exception of stimuli with rotations of $0^\prime$ or
$180^\prime$, in which case we ran double the number of samples). The
step size parameter ($\sigma_\epsilon$) was approximately fit to each
of the models, resulting in $\sigma_\epsilon=0.6$ for the \Bq{} model,
and $\sigma_\epsilon=0.1$ for all other models. We also ran the models
under two different priors, $p(h=0)=0.5$ (the ``equal'' prior) and
$p(h=0)=0.55$ (the ``unequal'' prior), though this only had an effect
for the \Bq{} model. \TODO{actually, did it affect HC?  check this}

@@

% The average response time across all stimuli was \ExpTime{} seconds.
% As expected, the minimum angle of rotation was significantly
% monotonically correlated with average per-stimulus response times,
% both for flipped pairs (\ExpThetaTimeCorrFlipped{}) and same pairs
% (\ExpThetaTimeCorrSame{}). This was also the case for the number of
% actions taken by the \Oc{} model (\OcThetaTimeCorr{}), the \Th{} model
% (\ThThetaTimeCorr{}), and the \Bq{} model (\BqThetaTimeCorr{}), but
% not the \Hc{} model (\HcThetaTimeCorr{}). The top row of
% Fig.~\ref{fig:response-time-accuracy} shows the overall effect of
% rotation on response times.

% The number of actions taken by the \Oc{} model was the best predictor
% of human response times, with a correlation of \ExpOcTimeCorr{}. The
% \Th{} model \Bq{} model with $p(h=0)=0.55$ were the next best (for
% \Th{}, \ExpThTimeCorr{}; for \Bq{}, \ExpBqpTimeCorr{}). Not far behind
% was the BQ model with equal prior (\ExpBqTimeCorr{}). The \Hc{} model
% was barely correlated with people (\ExpHcTimeCorr{}).

% Across all stimuli, participants' response times for ``flipped'' pairs
% were significantly higher than those for ``same'' pairs
% \TODO{confidence intervals} \TODO{flipped response times higher than
%   same response times for smaller angles?}

@@

% The average accuracy across all stimuli was \ExpAccuracy{}, though
% there were \ExpNumChance{} individual stimuli (out of 720) for which
% people were not above chance. Of the models, the \Oc{} and \Th{}
% models performed perfectly, with 100\% accuracy. The \Bq{} model was
% very accurate overall (equal prior: \BqAccuracy{}, unequal prior:
% \BqpAccuracy{}). With the equal prior, there was \BqNumChance{}
% stimulus for which it was not above chance; with the unequal prior,
% there were \BqpNumChance{}. The \Hc{} model was not particularly
% accurate, and although it did perform above chance overall
% (\HcAccuracy{}), there were \HcNumChance{} individual stimuli for
% which it was not above chance.

% The minimum angle was also correlated with participants' average
% per-stimulus accuracy: for flipped pairs,
% \ExpThetaAccuracyCorrFlipped{}, and for same pairs,
% \ExpThetaAccuracyCorrSame{}. There was also a significant effect for
% the \Hc{} model (\HcThetaAccuracyCorr{}) and the \Bq{} model (equal
% prior: \BqThetaAccuracyCorr{}, unequal prior:
% \BqpThetaAccuracyCorr{}). The bottom row of
% Fig.~\ref{fig:response-time-accuracy} shows the effect of angle on
% accuracy for both participants and models.

% There was a moderate correlation between the accuracy of the \Hc{}
% model and human accuracy (\ExpHcAccuracyCorr{}). The \Bq{} model had a
% similar correlation (equal prior: \ExpBqAccuracyCorr{}, unequal prior:
% \ExpBqpAccuracyCorr{}). Correlations for the other models were
% undefined, because they had 100\% accuracy.

\begin{figure*}[t]
  \begin{center}
    \subfloat{\includegraphics[width=0.33\textwidth]{../../figures/D/theta_time_corrs.pdf}}
    \subfloat{\includegraphics[width=0.33\textwidth]{../../figures/D/theta_accuracy_corrs.pdf}}
    \subfloat{\includegraphics[width=0.33\textwidth]{../../figures/D/response_time_corrs.pdf}}
    \caption{}
    \label{fig:response-time-corrs}
  \end{center}
\end{figure*}

@@

We replicated the previous monotonic response time result, though not
the linear result. Actually kind of similar to \cite{Gardony:2013gn}
\TODO{elaborate}
